---
output: 
  html_document:
    toc: TRUE
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


# Introduction to RMarkdown

RMarkdown is a powerful "literate programming" environment, meaning that one can easily combine text, code and output - including interactive `htmlwidgets`. There are three types of document that can be created:

- PDF: Useful when you want to print documents (and remains the default output format in academic publishing)
- MS Word: Useful if others you're working with **need** to work with MS Word documents
- HTML: Useful for creating interactive content you can publish to the web

All HTML documents (including reports and presentations) created with RMarkdown can be published freely (and publicly) to [RPubs.com](RPubs.com).

All of the content below uses the same code as from the `cheats_data-viz.R` file but demonstrates how different types of content can be combined. The data behind these examples was extrapolated and anonymised from the following Figshare deposit: [https://doi.org/10.6084/m9.figshare.4516772](https://doi.org/10.6084/m9.figshare.4516772).

The following packages are used in this document, and all data is from the `data/cheats_journeys.csv` file.

```{r echo=TRUE}
library("tidyverse")
library("leaflet")
library("highcharter")
library("sf")
library("statesRcontiguous")
library("lubridate")
journeys <- read_csv("data/cheats_journeys.csv")
```


# Maps

The leaflet library allows us to create interactive maps, here are two common GIS visualsiations. To learn morw about GIS visualisations, please do visit the [Oxshef Charts website](https://oxshef.github.io/oxshef_charts/geoviz_maps.html).

## Scatter geo plots

scatter geo plots are very useful for simply demonstrating *where* events occured. In the map below circle markers are used to show the start and end locations in red and green, respectively.

```{r leaflet-scatter-geoplot}
journeys %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~ start.longitude,
    lat = ~ start.latitude,
    color = "red",
    radius = 1
  ) %>%
  addCircleMarkers(
    lng = ~ end.longitude,
    lat = ~ end.latitude,
    color = "green",
    radius = 1
  )
```

## Choropleth

The `leaflet` library allows us to easily visualise choropleth, provided we have appropriate shapefiles available. Here we prepare a shaepfile of the contiguous USA using the `statesRcontiguous` library:

```{r echo=TRUE}
contiguous_usa <- shp_all_us_states %>%
  filter(contiguous.united.states == TRUE)
```

Using the `sf` library it is fairly easy to calculate the number of journeys that started in each State:

```{r echo=TRUE}
state_send_locs <- journeys %>%
  filter(start.country == "USA") %>%
  st_as_sf(
    coords = c("start.longitude", "start.latitude"),
    crs = st_crs(contiguous_usa)
  )

contiguous_send_counts <- contiguous_usa %>%
  mutate(send.counts = st_covers(contiguous_usa, state_send_locs) %>%
           lengths())
```

Finally, we can create a palette and visualise this data:

```{r echo=TRUE}
palette_contiguous_us <-
  colorBin("YlOrBr", domain = contiguous_send_counts$send.counts)

contiguous_send_counts %>%
  leaflet() %>%
  addPolygons(
    fillColor = ~ palette_contiguous_us(send.counts),
    fillOpacity = 1,
    weight = 1,
    color = "#000",
    label = ~ paste0(state.name, " (", send.counts, " journeys)") 
  ) %>%
  addLegend(pal = palette_contiguous_us,
            values = ~send.counts)
```

# Calendar heatmap

`ggplot2` is an extraudinarily powerful and versatile tool for visualising data with R, implementing a consistent and complete "grammar of graphics". As an example of how far you can go with `ggplot2` this is a calendar heatmap showing the distribution of letters sent through the 1860s in the journeys dataset.

```{r}
dated_journeys <- journeys %>%
  select(date, number.of.letters) %>%
  mutate(year = year(date),
         yearmonthf = paste(month(date, label = TRUE), year(date)),
         monthf = month(date, abbr = TRUE, label = TRUE),
         week = week(date),
         monthweek = ceiling(day(journeys$date) / 7),
         weekdayf = wday(date,label = TRUE))

gg_dated_journeys <- dated_journeys %>%
  filter(date >= dmy("01-01-1860") & date <= dmy("01-12-1869")) %>%
  ggplot(aes(monthweek, weekdayf, fill = number.of.letters)) + 
  geom_tile(colour = "white") + 
  facet_grid(year~monthf) +
  scale_fill_gradient(low="red", high="green") +
  labs(x="Week of Month",
       y="",
       title = "Number of letters sent per day in the 1860s", 
       fill = "Number of letters")
gg_dated_journeys
```

# Comparing ggplot2 and highcharter

The `highcharter` library allows us to create very professional looking interactive charts and plots. It's important to note that the library is **NOT** free to use for commercial or governmental usage, though we can use it when communicating research outputs.

Both `ggplot2` and `highcharter` use a very similar syntax:

- `ggplot(data, aes(x = x, y = y))`: the `aes` function sets the aesthetics for the ggplot object, which columns in `data` should be used for which visual properties of the chart.
- `hchart(data, hcaes(x = x, y = y))`: the `hcaes` function sets the aesthetics for the highcharter object, which columns in `data` should be used for which visual properties of the chart

## Country -> Country journeys tally {.tabset}

If you refer to the .Rmd file used to generate this report, you'll notice the section header has {.tabset} appended. This allows us to create the tabbed content below from any child subheading of the current heading level.

### ggplot2

```{r}
country_to_country_counts <- journeys %>%
  count(start.country, end.country) %>%
  mutate(journey = paste(start.country, "->", end.country)) %>%
  arrange(n) %>%
  mutate(journey = as.factor(journey)) %>%
  mutate(journey = fct_reorder(journey, n))

gg_country_to_country_counts <- country_to_country_counts %>%
  ggplot(aes(x = journey, y = n)) + geom_col() +
  coord_flip() +
  xlab("") +
  ylab("Number of journeys") +
  ggtitle("Number of journeys split by start and end country")
gg_country_to_country_counts
```

### highcharter

```{r}
journeys %>%
  count(start.country, end.country) %>%
  mutate(journey = paste(start.country, "->", end.country)) %>%
  arrange(desc(n)) %>%
  hchart(type = "bar",
         hcaes(x = journey, y = n)) %>%
  hc_xAxis(title = list(text = "")) %>%
  hc_yAxis(title = list(text = "Number of journeys")) %>%
  hc_title(text = "Number of journeys split by start and end country")
```

## Grouped bar charts {.tabset}

When constructing grouped or stacked bar charts in `ggplot2` or `highcharter` one must ensure to reshape data into long format:

- variable column: a measure of the data 
- value column: a value for a specific measure of the data

This is achieved with `gather` from the `tidyr` library.

```{r}
end_country_tallies <- journeys %>%
  group_by(end.country) %>%
  summarise(total.letters = sum(number.of.letters),
            total.journeys = n()) %>%
  mutate(
    total.letters = total.letters / sum(total.letters),
    total.journeys = total.journeys / sum(total.journeys)
  ) %>%
  arrange(total.letters) %>%
  mutate(end.country = fct_reorder(end.country, total.letters)) %>%
  gather(measure, value,-end.country) 
end_country_tallies
```

### ggplot2

```{r}
end_country_tallies %>%
  ggplot(aes(x = end.country,
             y = value,
             fill = measure)) +
  geom_col(position = "dodge") +
  coord_flip() +
  xlab("Final Destination Country") +
  ylab("Percentage") +
  ggtitle("Final destination of letters",
          subtitle = "Ordered by percentage of journeys") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(
    values = c("#1b9e77", "#7570b3"),
    name = "",
    breaks = c("total.letters", "total.journeys"),
    labels = c("Percentage of Journeys", "Percentage of letters")
  )
```

### highcharter

```{r}
end_country_tallies %>%
  hchart(type = "bar",
         hcaes(x = end.country,
               y = 100 * value,
               group = measure)) %>%
  hc_xAxis(title = list(text = "Final Destination Country"),
           reversed = FALSE) %>%
  hc_yAxis(title = list(text = "Percentage"),
           labels = list(format = '{value}%')) %>%
  hc_title(text = "Final destination of letters") %>%
  hc_subtitle(text = "Ordered by percentage of journeys")
```

